To Batch or Not to Batch: Test-Ordering Variability in the ED

### 2024-09-09 manuscript/CopyOfmanuscript.tex
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits...
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
---

# Introduction {#sec:I}

Healthcare delivery, particularly in the emergency department (ED), requires a delicate balance between ensuring optimal patient outcomes and optimizing resource utilization. Achieving these twin goals necessitates timely and accurate diagnosis, which enables prompt and appropriate treatment, improving patient prognosis and reducing the likelihood of adverse events. Efficient patient discharge from the ED is crucial for alleviating overcrowding, a severe issue associated with higher complication rates and increased mortality @bernstein2009. In this complex environment, the availability and performance of diagnostic tests, especially imaging studies, play a pivotal role in shaping both patient care and operational efficiency @naseim2015.

A critical question in ED management pertains to whether physicians should batch order diagnostic imaging tests or order them sequentially. This decision represents a fundamental tradeoff between potentially reducing patient length of stay and risking over-testing. Batch ordering, where multiple imaging tests are ordered simultaneously, may expedite the diagnostic process but can lead to unnecessary tests, increased costs, patient anxiety, and potential harm from follow-up of false-positive results @koch2018. Conversely, a more sequential approach might reduce unnecessary testing but could extend a patient's time in the ED, potentially contributing to overcrowding. Striking the right balance between comprehensive evaluation and efficient resource use is crucial, yet research on optimal test ordering strategies in the ED remains limited @saghafian2015.

@@

To address potential confounding factors, we employ a placebo exercise to demonstrate that the observed effects are primarily due to batching behavior rather than other aspects of physician care that might correlate with batching tendency. We find no significant effects of physician batch tendency on key outcomes for a "placebo sample" of patients visiting the ED for conditions that rarely result in batched orders. This strengthens our confidence in the specificity of the batching effect on patient outcomes and resource utilization.

@@

# Theoretical Model

@@

In a typical ED scenario, patients arrive with varying levels of acuity and a wide range of presenting symptoms. Upon arrival, each patient $i$ is assigned to a physician $j$, often based on availability rather than specific matching criteria. The physician's primary task is to diagnose and treat the patient efficiently and effectively, a process that frequently involves ordering diagnostic tests, including imaging studies.

Central to our model is the concept of the patient's underlying condition, denoted by $\tau_i$. This represents the true state of the patient's health, which is initially unknown to the treating physician. The physician's goal is to uncover $\tau_i$ as accurately and quickly as possible to provide appropriate care. However, the path to this discovery is not straightforward and involves critical decisions about diagnostic testing.

The physician's approach to ordering tests is influenced by various factors, including their individual tendencies and experiences. We represent this as $Z_j$, the physician's propensity to batch order tests. This tendency can be thought of as a stable characteristic of the physician, shaped by their training, past experiences, and personal practice style. The decision to batch order tests for a given patient, represented by $X_{ij}$, is a function of both the physician's tendency $Z_j$ and the specifics of the patient's presentation. Batch ordering involves requesting multiple imaging tests simultaneously, rather than in a sequential manner. This decision has cascading effects on the patient's journey through the ED.

@@

As illustrated in Figure \ref{fig:dag}, when a physician chooses to batch order ($X_{ij} = 1$), it directly influences the total volume of imaging tests performed, $M_{ij}$. Our model posits two key mechanisms for this relationship:

@@

The total volume of tests, $M_{ij}$, in turn, impacts key patient outcomes, $Y_{ij}$, particularly length of stay (LOS) and the probability of a 72-hour return visit with admission. The relationship between test volume and these outcomes is complex. On one hand, more tests generally require more time, potentially increasing LOS. However, if the physician's initial diagnostic uncertainty is high, ordering more tests upfront might expedite the diagnostic process and lead to quicker treatment decisions.

Importantly, our model also allows for a direct effect of the batch ordering decision on outcomes, independent of its effect through test volume. This direct effect could represent, for example, the efficiency gains from parallel processing of multiple tests, which might reduce LOS even as the total number of tests increases. The model acknowledges the presence of unobserved confounding factors, $U_{ij}$, which might influence both the decision to batch order and patient outcomes. These could include factors like the physician's assessment of the patient's condition severity or subtle clinical cues not captured in standard data.

@@

To address potential endogeneity, our empirical strategy leverages the physician's general tendency to batch order, $Z_j$, as an instrument for the actual batching decision, $X_{ij}$. This approach allows us to estimate the causal effects of batch ordering on test volume and patient outcomes, even in the presence of unobserved confounders.

@@

In summary, our theoretical model provides a framework for understanding the complex decision-making process in ED diagnostic testing. It highlights the potential trade-offs involved in batch ordering – between comprehensive initial assessment and efficient resource use, between rapid testing and informed sequential decision-making. By formalizing these relationships, we set the stage for our empirical analysis, which aims to quantify these effects and provide insights for optimizing ED operations and patient care.

# Data and Definitions {#sec:II}

Our study was conducted in the Emergency Department (ED) of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation. During the study period, the ED recorded approximately 43,000 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces. The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume. Physicians operate in a unique workflow that includes staggered 8.5-hour shifts and a rotational patient assignment system, minimizing potential selection bias in patient encounters.

@@

## Sample Construction and Summary Statistics

Our research design focuses on adult visits to the Mayo Clinic of Arizona ED. From an initial dataset of approximately 48,000 visits during the study period, we applied several exclusion criteria to improve statistical power and ensure the validity of our physician batch tendency instrument. We excluded encounters with rare chief complaints (fewer than 1000 total encounters) and complaints where batch ordering occurs in less than 5% of cases. Examples of excluded complaints include Upper Respiratory Symptoms and Urinary Complaints. This exclusion strategy is unlikely to introduce selection bias as long as physician test batching tendency is not strongly correlated with diagnosing behavior across broad complaint categories. To estimate a precise measure of physician-level batch tendency, we further restricted our sample to encounters involving full-time physicians who treat over 500 ED cases per year. These criteria resulted in a final analytical sample of 16,361 patient encounters.

Table \ref{tab:summary_statistics} provides a comprehensive overview of ED characteristics, patient demographics, and medical tests in our study site. The ED manages an average volume of 24.3 patients, indicating a significant yet manageable workload. The patient population exhibits a range of critical conditions, with 18.8% presenting as tachycardic, 9.51% as tachypneic, 2.43% with fever, and 1.57% as hypotensive. Diagnostic procedures in our sample heavily rely on laboratory tests, used in 78.5% of encounters. Imaging plays a crucial role, with X-rays ordered in 48.1% of visits, non-contrast CT scans in 23.1%, and contrast CT scans in 20.5%. Ultrasounds are used in 12.5% of encounters. Notably, 6.32% of visits involve batch ordering of imaging tests, a key focus of our study.

@@

## Variable Definitions

Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter. While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice. Below, we detail our primary outcomes: (a) ED length of stay (LOS), and (b) total imaging volume, and (c) 72-hour return with admission.

*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings. It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital. The hypothesis is that batch testing may lead to shorter ED-LOS, potentially improving patient throughput and reducing crowding.

*(b) Resource Utilization*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives. In this study, we quantify resource utilization by the total number of distinct diagnostic tests ordered per patient during their ED stay. This encompasses both initial and subsequent tests. The hypothesis is that batch testing may lead to variations in the number of tests ordered, potentially influencing the overall healthcare expenditure and efficiency.

*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital. This outcome is used to assess the quality of care provided during the initial ED visit. The hypothesis is that batch testing may lead to variations in the quality of care provided, potentially influencing the likelihood of a return visit.

### Batching

We define "batching" in line with standard emergency medicine practices. Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses. This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.

@@

# Empirical Strategy {#sec:3}

Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court. Similarly, we explore batching variation across physicians who work in the same emergency department. In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians. Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision. We employ both approaches and lay out their details in the next subsections.

## Institutional Details on Patient-Physician Assignment

Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED. In most EDs, however, physicians have discretion in picking their patients. In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses. In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random. Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.

@@

## Batch Tendency Construction

To measure physician batch tendency, we use the physician's residualized leave-out average batch rate. This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways. First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.

@@

where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$. This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument. The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.

@@

\hyperref[table:first_stage]{Table } illustrates the "first stage" results, emphasizing the robust relevance of the instrument across various controls. Being assigned to a $10$pp higher batch-tendency physician is associated with a $17.6$pp increase in the probability of having tests batch-ordered in the ED. The F-statistic is 21.6 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only.

@@

Where $Tests_i$ is the number of imaging tests ordered, and $Y_i$ represents our patient outcomes of interest: length of stay and 72-hour return with admission. $X_i$ is the same set of control variables as in the reduced-form approach. The $Batched_i$ variable suffers from potential endogeneity concerns. For example, unobserved injury severity may correlate with both the need for multiple tests and patient outcomes. To address this, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We use a causal mediation analysis to decompose the total effect of batching into its direct effect on outcomes and its indirect effect through the number of tests ordered. This approach employs bootstrapping to estimate confidence intervals for the direct and indirect effects. We cluster standard errors at the physician level to account for potential correlation in outcomes among patients treated by the same physician, and to properly adjust for the shared physician-level instrument across patients.

## Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence"). The residualization in equation (4) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.

@@

Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects. It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician. The literature leveraging the judges design typically performs two informal tests for its implications. The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects). The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration). Appendix Table  \ref{tab:monotonicity} presents both of these tests in the two columns for various subsamples of interest.

# Results {#sec:4}

## Reduced-Form Results

@@

## Placebo Check

In this section we investigate whether the reduced-form effects observed in Section 5.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency. We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing. For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes. Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.

To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a $>5$ percent batching rate). We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (6). The results of this exercise are displayed in Appendix Table 1. They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.

## Instrumental Variables Results

@@

### External Validity

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[,,nonblindrev]{informs}

\OneAndAHalfSpacedXI
%%\OneAndAHalfSpacedXII % Current default line spacing
%%\DoubleSpacedXII
%%\DoubleSpacedXI

%% BEGIN MY ADDITIONS %%
\usepackage{hyperref}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}
\usepackage{caption}

%% END MY ADDITIONS %%

% Natbib setup for author-year style
\usepackage{natbib}
 \bibpunct[, ]{(}{)}{,}{a}{}{,}%
 \def\bibfont{\small}%
 \def\bibsep{\smallskipamount}%
 \def\bibhang{24pt}%
 \def\newblock{\ }%
 \def\BIBand{and}%

%% Setup of theorem styles. Outcomment only one.
%% Preferred default is the first option.
\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
\ECRepeatTheorems

%% Setup of the equation numbering system. Outcomment only one.
%% Preferred default is the first option.
\EquationsNumberedThrough    % Default: (1), (2), ...
%\EquationsNumberedBySection % (1.1), (1.2), ...

% For new submissions, leave this number blank.
% For revisions, input the manuscript number assigned by the on-line
% system along with a suffix ".Rx" where x is the revision number.
\MANUSCRIPTNO{}

%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%

% Outcomment only when entries are known. Otherwise leave as is and
%   default values will be used.
%\setcounter{page}{1}
%\VOLUME{00}%
%\NO{0}%
%\MONTH{Xxxxx}% (month or a similar seasonal id)
%\YEAR{0000}% e.g., 2005
%\FIRSTPAGE{000}%
%\LASTPAGE{000}%
%\SHORTYEAR{00}% shortened year (two-digit)
%\ISSUE{0000} %
%\LONGFIRSTPAGE{0001} %
%\DOI{10.1287/xxxx.0000.0000}%

% Author's names for the running heads
% Sample depending on the number of authors;
\RUNAUTHOR{%
Jameson, Saghafian, Huckman
 and Hodgson
}
% \RUNAUTHOR{Jones and Wilson}
% \RUNAUTHOR{Jones, Miller, and Wilson}
% \RUNAUTHOR{Jones et al.} % for four or more authors
% Enter authors following the given pattern:
%\RUNAUTHOR{}

\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}

\TITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}

\ARTICLEAUTHORS{%
\AUTHOR{Jacob Jameson}
\AFF{Interfaculty Initiative in Health Policy, Harvard
University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}

\AUTHOR{Soroush Saghafian}
\AFF{Kennedy School of Government, Harvard
University, \EMAIL{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}}}

\AUTHOR{Robert Huckman}
\AFF{Harvard Business
School, \EMAIL{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}}}

\AUTHOR{Nicole Hodgson}
\AFF{Mayo
Clinic, \EMAIL{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}}}

%
}

\ABSTRACT{We use practice variation across physicians to uncover the
role of test-ordering on care delivery in the ED. Using records of over
45,000 Mayo Clinic emergency department visits\ldots{}}

\KEYWORDS{Emergency Department, Operational Effeciency, Diagnostic
Testing}

\maketitle

\hypertarget{sec:I}{%
\section{Introduction}\label{sec:I}}

Healthcare delivery, particularly in the emergency department (ED),
requires a delicate balance between ensuring optimal patient outcomes
and optimizing resource utilization. Achieving these twin goals
necessitates timely and accurate diagnosis, which enables prompt and
appropriate treatment, improving patient prognosis and reducing the
likelihood of adverse events. Efficient patient discharge from the ED is
crucial for alleviating overcrowding, a severe issue associated with
higher complication rates and increased mortality \citet{bernstein2009}.
In this complex environment, the availability and performance of
diagnostic tests, especially imaging studies, play a pivotal role in
shaping both patient care and operational efficiency \citet{naseim2015}.

A critical question in ED management pertains to whether physicians
should batch order diagnostic imaging tests or order them sequentially.
This decision represents a fundamental tradeoff between potentially
reducing patient length of stay and risking over-testing. Batch
ordering, where multiple imaging tests are ordered simultaneously, may
expedite the diagnostic process but can lead to unnecessary tests,
increased costs, patient anxiety, and potential harm from follow-up of
false-positive results \citet{koch2018}. Conversely, a more sequential
approach might reduce unnecessary testing but could extend a patient's
time in the ED, potentially contributing to overcrowding. Striking the
right balance between comprehensive evaluation and efficient resource
use is crucial, yet research on optimal test ordering strategies in the
ED remains limited \citet{saghafian2015}.

@@

To address potential confounding factors, we employ a placebo exercise
to demonstrate that the observed effects are primarily due to batching
behavior rather than other aspects of physician care that might
correlate with batching tendency. We find no significant effects of
physician batch tendency on key outcomes for a ``placebo sample'' of
patients visiting the ED for conditions that rarely result in batched
orders. This strengthens our confidence in the specificity of the
batching effect on patient outcomes and resource utilization.

@@

\hypertarget{theoretical-model}{%
\section{Theoretical Model}\label{theoretical-model}}

@@

In a typical ED scenario, patients arrive with varying levels of acuity
and a wide range of presenting symptoms. Upon arrival, each patient
\(i\) is assigned to a physician \(j\), often based on availability
rather than specific matching criteria. The physician's primary task is
to diagnose and treat the patient efficiently and effectively, a process
that frequently involves ordering diagnostic tests, including imaging
studies.

Central to our model is the concept of the patient's underlying
condition, denoted by \(\tau_i\). This represents the true state of the
patient's health, which is initially unknown to the treating physician.
The physician's goal is to uncover \(\tau_i\) as accurately and quickly
as possible to provide appropriate care. However, the path to this
discovery is not straightforward and involves critical decisions about
diagnostic testing.

The physician's approach to ordering tests is influenced by various
factors, including their individual tendencies and experiences. We
represent this as \(Z_j\), the physician's propensity to batch order
tests. This tendency can be thought of as a stable characteristic of the
physician, shaped by their training, past experiences, and personal
practice style. The decision to batch order tests for a given patient,
represented by \(X_{ij}\), is a function of both the physician's
tendency \(Z_j\) and the specifics of the patient's presentation. Batch
ordering involves requesting multiple imaging tests simultaneously,
rather than in a sequential manner. This decision has cascading effects
on the patient's journey through the ED.

@@

As illustrated in Figure \ref{fig:dag}, when a physician chooses to
batch order (\(X_{ij} = 1\)), it directly influences the total volume of
imaging tests performed, \(M_{ij}\). Our model posits two key mechanisms
for this relationship:

@@

The total volume of tests, \(M_{ij}\), in turn, impacts key patient
outcomes, \(Y_{ij}\), particularly length of stay (LOS) and the
probability of a 72-hour return visit with admission. The relationship
between test volume and these outcomes is complex. On one hand, more
tests generally require more time, potentially increasing LOS. However,
if the physician's initial diagnostic uncertainty is high, ordering more
tests upfront might expedite the diagnostic process and lead to quicker
treatment decisions.

Importantly, our model also allows for a direct effect of the batch
ordering decision on outcomes, independent of its effect through test
volume. This direct effect could represent, for example, the efficiency
gains from parallel processing of multiple tests, which might reduce LOS
even as the total number of tests increases. The model acknowledges the
presence of unobserved confounding factors, \(U_{ij}\), which might
influence both the decision to batch order and patient outcomes. These
could include factors like the physician's assessment of the patient's
condition severity or subtle clinical cues not captured in standard
data.

@@

To address potential endogeneity, our empirical strategy leverages the
physician's general tendency to batch order, \(Z_j\), as an instrument
for the actual batching decision, \(X_{ij}\). This approach allows us to
estimate the causal effects of batch ordering on test volume and patient
outcomes, even in the presence of unobserved confounders.

@@

In summary, our theoretical model provides a framework for understanding
the complex decision-making process in ED diagnostic testing. It
highlights the potential trade-offs involved in batch ordering --
between comprehensive initial assessment and efficient resource use,
between rapid testing and informed sequential decision-making. By
formalizing these relationships, we set the stage for our empirical
analysis, which aims to quantify these effects and provide insights for
optimizing ED operations and patient care.

\hypertarget{sec:II}{%
\section{Data and Definitions}\label{sec:II}}

Our study was conducted in the Emergency Department (ED) of the Mayo
Clinic of Arizona, a tertiary care hospital without obstetrical
services, an inpatient pediatrics unit, or a trauma designation. During
the study period, the ED recorded approximately 43,000 visits per year,
managed across 26 treatment rooms and up to 9 hallway spaces. The
department is exclusively staffed by board-eligible or board-certified
emergency physicians (EPs), with rotating residents overseeing about
10\% of patient volume. Physicians operate in a unique workflow that
includes staggered 8.5-hour shifts and a rotational patient assignment
system, minimizing potential selection bias in patient encounters.

@@

\hypertarget{sample-construction-and-summary-statistics}{%
\subsection{Sample Construction and Summary
Statistics}\label{sample-construction-and-summary-statistics}}

Our research design focuses on adult visits to the Mayo Clinic of
Arizona ED. From an initial dataset of approximately 48,000 visits
during the study period, we applied several exclusion criteria to
improve statistical power and ensure the validity of our physician batch
tendency instrument. We excluded encounters with rare chief complaints
(fewer than 1000 total encounters) and complaints where batch ordering
occurs in less than 5\% of cases. Examples of excluded complaints
include Upper Respiratory Symptoms and Urinary Complaints. This
exclusion strategy is unlikely to introduce selection bias as long as
physician test batching tendency is not strongly correlated with
diagnosing behavior across broad complaint categories. To estimate a
precise measure of physician-level batch tendency, we further restricted
our sample to encounters involving full-time physicians who treat over
500 ED cases per year. These criteria resulted in a final analytical
sample of 16,361 patient encounters.

Table \ref{tab:summary_statistics} provides a comprehensive overview of
ED characteristics, patient demographics, and medical tests in our study
site. The ED manages an average volume of 24.3 patients, indicating a
significant yet manageable workload. The patient population exhibits a
range of critical conditions, with 18.8\% presenting as tachycardic,
9.51\% as tachypneic, 2.43\% with fever, and 1.57\% as hypotensive.
Diagnostic procedures in our sample heavily rely on laboratory tests,
used in 78.5\% of encounters. Imaging plays a crucial role, with X-rays
ordered in 48.1\% of visits, non-contrast CT scans in 23.1\%, and
contrast CT scans in 20.5\%. Ultrasounds are used in 12.5\% of
encounters. Notably, 6.32\% of visits involve batch ordering of imaging
tests, a key focus of our study.

@@

\hypertarget{variable-definitions}{%
\subsection{Variable Definitions}\label{variable-definitions}}

Our explanatory variable in the IV analysis, \(Batched_i\), is an
indicator for whether patient \(i\) has their tests batch-ordered at
their ED encounter. While the patient could decide not to undergo the
tests ordered by the physician, this is rare in practice. Below, we
detail our primary outcomes: (a) ED length of stay (LOS), and (b) total
imaging volume, and (c) 72-hour return with admission.

\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
efficiency and patient throughput in emergency care settings. It is
defined as the duration from a patient's arrival to the ED until their
departure, whether by discharge or admission to the hospital. The
hypothesis is that batch testing may lead to shorter ED-LOS, potentially
improving patient throughput and reducing crowding.

\emph{(b) Resource Utilization}-- Resource utilization in the ED
typically refers to the extent of medical services and interventions a
patient receives. In this study, we quantify resource utilization by the
total number of distinct diagnostic tests ordered per patient during
their ED stay. This encompasses both initial and subsequent tests. The
hypothesis is that batch testing may lead to variations in the number of
tests ordered, potentially influencing the overall healthcare
expenditure and efficiency.

\emph{(c) 72 Hour Return with Admission}-- This is a binary variable
indicating whether the patient returns to the ED within 72 hours of
their initial visit and is admitted to the hospital. This outcome is
used to assess the quality of care provided during the initial ED visit.
The hypothesis is that batch testing may lead to variations in the
quality of care provided, potentially influencing the likelihood of a
return visit.

\hypertarget{batching}{%
\subsubsection{Batching}\label{batching}}

We define ``batching'' in line with standard emergency medicine
practices. Batching occurs when a physician simultaneously orders a
comprehensive set of diagnostic tests, typically covering a broad range
of potential diagnoses. This contrasts with sequential ordering, where
tests are ordered sequentially based on the information obtained from
each test as needed.

@@

\hypertarget{sec:3}{%
\section{Empirical Strategy}\label{sec:3}}

Our empirical strategy closely follows the literature that relies on
quasi-random assignment of agents to cases, often referred to as the
``judges design.'' Papers in this literature typically exploit variation
in the sentencing leniency of judges who work in the same court.
Similarly, we explore batching variation across physicians who work in
the same emergency department. In its reduced form, under the assumption
of quasi-random assignment, this approach allows researchers to identify
the causal effect of being assigned to different types of physicians.
Under additional assumptions, an instrumental variable approach
identifies the causal effect of a given medical decision. We employ both
approaches and lay out their details in the next subsections.

\hypertarget{institutional-details-on-patient-physician-assignment}{%
\subsection{Institutional Details on Patient-Physician
Assignment}\label{institutional-details-on-patient-physician-assignment}}

Contrary to most healthcare settings where patients exhibit choice, they
are predominantly passive in their physician assignment in the ED. In
most EDs, however, physicians have discretion in picking their patients.
In contrast, patients arriving at the Mayo Clinic ED are randomly
assigned to physicians via a rotational patient assignment algorithm
\citet{traub2016emergency}, which removes potential selection bias
concerns for our analyses. In essence, barring arrival time and
shift-level variation, the physician-to-patient matching can be deemed
random. Table \ref{tab:wald_test} displays that patient encounters
(regarding chief complaints and emergency severity) are equitably
distributed across physicians within our study's cohort.

@@

\hypertarget{batch-tendency-construction}{%
\subsection{Batch Tendency
Construction}\label{batch-tendency-construction}}

To measure physician batch tendency, we use the physician's residualized
leave-out average batch rate. This measure is derived from two steps
following the approaches taken by \citet{doyle2015measuring},
\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
obtain residuals from a regression model, which includes all ED
encounters in our sample period.

@@

where \(\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}\) is the
residual from equation (1); \(J\) is the set of all ED encounters
treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
number of cases that physician has seen that year, excluding patient
\(i\). This leave-out mean eliminates the mechanical bias that stems
from patient \(i\)'s own case entering into the instrument. The measure
is interpreted as the average (leave-out) batch rate of patient \(i\)'s
physician, relative to other physicians in that hospital-year-month,
hospital-day of week-time of day.

@@

\hyperref[table:first_stage]{Table } illustrates the ``first stage''
results, emphasizing the robust relevance of the instrument across
various controls. Being assigned to a \(10\)pp higher batch-tendency
physician is associated with a \(17.6\)pp increase in the probability of
having tests batch-ordered in the ED. The F-statistic is 21.6 when all
controls and fixed effects are included. The coefficient is greater than
one because all emergency visits are used to construct the tendency
instrument, while the first stage is calculated using the baseline
sample only.

@@

Where \(Tests_i\) is the number of imaging tests ordered, and \(Y_i\)
represents our patient outcomes of interest: length of stay and 72-hour
return with admission. \(X_i\) is the same set of control variables as
in the reduced-form approach. The \(Batched_i\) variable suffers from
potential endogeneity concerns. For example, unobserved injury severity
may correlate with both the need for multiple tests and patient
outcomes. To address this, we instrument \(Batched_i\) with the assigned
physician \(j\)'s underlying tendency to batch,
\(Tendency_{i,j}^{phys}\). We use a causal mediation analysis to
decompose the total effect of batching into its direct effect on
outcomes and its indirect effect through the number of tests ordered.
This approach employs bootstrapping to estimate confidence intervals for
the direct and indirect effects. We cluster standard errors at the
physician level to account for potential correlation in outcomes among
patients treated by the same physician, and to properly adjust for the
shared physician-level instrument across patients.

\hypertarget{identifying-assumptions}{%
\subsection{Identifying Assumptions}\label{identifying-assumptions}}

The reduced-form approach delivers an unbiased estimate of the causal
effect of being treated by a higher tendency to batch physician, since
assignment of patients to ED physicians is random, conditional on
seasonality and shift (``conditional independence''). The
residualization in equation (4) controls for more controls than required
to achieve quasi-random assignment; they are included for statistical
precision in measuring physician tendency to batch.

@@

Finally, the monotonicity assumption is necessary for interpreting the
coefficient estimates obtained from the IV approach as Local Average
Treatment Effects (LATEs) if there are heterogeneous treatment effects.
It requires that any patient who is (not) batched by a sequencer
(batcher) would also (not) be batched by a batcher (sequencer)
physician. The literature leveraging the judges design typically
performs two informal tests for its implications. The first one provides
that the first stage should be weakly positive for all subsamples
(\citet{dobbie2018effects}). The second implication asserts that the
instrument constructed by leaving out a particular subsample has
predictive power over that same left-out subsample
(\citet{bhuller2020incarceration}). Appendix Table
\ref{tab:monotonicity} presents both of these tests in the two columns
for various subsamples of interest.

\hypertarget{sec:4}{%
\section{Results}\label{sec:4}}

\hypertarget{reduced-form-results}{%
\subsection{Reduced-Form Results}\label{reduced-form-results}}

@@

\hypertarget{placebo-check}{%
\subsection{Placebo Check}\label{placebo-check}}

In this section we investigate whether the reduced-form effects observed
in Section 5.1 are due to differences in batch rates across providers or
due to other provider differences correlated with batch tendency. We
start by studying reduced-form effects among patients with complaints
that are never batched, as a ``placebo/falsification check.'' By way of
example, consider a patient who arrives at the ED with a urinary tact
infection---a condition for which patients rarely undergo imaging
testing. For such patients, we should expect to see no impact of batch
tendency only if high batching and low batching physicians do not
systematically differ in other dimensions of care relevant to patient
outcomes. Conversely, if we do find a reduced-form effect for these
patients, then high batch tendency physicians must systematically differ
from low batch tendency physicians in other dimensions of care, beyond
batching.

To that end, we restrict attention to ED visits for complaints where
batching occurs no more than 5 percent of the time (recall that our
baseline sample only includes complaints with a \(>5\) percent batching
rate). We estimate a reduced-form regression of each main outcome on
physician batch tendency for the subsample, following equation (6). The
results of this exercise are displayed in Appendix Table 1. They show
that in contrast to results for our main sample, the association between
physician tendency to batch and a given outcome is statistically
indistinguishable from zero and much smaller in magnitude for the
samples of patients who visit the ED with health conditions that are
rarely batched.

\hypertarget{instrumental-variables-results}{%
\subsection{Instrumental Variables
Results}\label{instrumental-variables-results}}

@@

\hypertarget{external-validity}{%
\subsubsection{External Validity}\label{external-validity}}
